---
title: "Activity Monitoring Data"
author: "JFoo"
date: "6/8/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and Preprocessing The Data

The data is loaded and stored into variable `Data` using `read.csv()` function. Then the `date` column is converted  from `character` vector into `Date` vector.  
```{r Loading & Preprocessing}
Data <- read.csv("activity.csv")
Data$date <- as.Date(Data$date, format = "%Y-%m-%d")
```  
  
  
##  Total Number of Steps Taken Per Day 

The total number of steps taken per day is calculated, ignoring missing value in dataset. Using  `tapply()` function, the total of steps is grouped by day and the sum is calculated.   
```{r Total Steps Per Day}
byDay <- tapply(Data$steps, Data$date, sum, na.rm=TRUE)
print(byDay)
```

The histogram of the total number of steps taken each day using is plotted using `ggplot`. Note the correpondence between value calculated above and the histogram plotted below.

```{r Histogram_for_total_steps_per_day}
library(ggplot2)
ggplot(Data) + geom_histogram(aes(x=date, weight=steps), bins=length(unique(Data$date))) + 
    labs(x="Date", y="No of Steps", title = "Histogram of The Total Number of Steps Taken Each Day")
```

Finally, the mean and median of total number of steps taken per day is calculated using functions as shown below: 

```{r Mean and Median}
mean <- mean(byDay)
median <- median(byDay)
```

Thus, we founded that the mean is `r mean` and the medial is `r median`.  


## Average Daily Activity Pattern

To calculate the average number of steps taken across all day for every 5 minutes interval, the data is grouped by interval and the mean of each interval is then calculated. `tapply()` function is used to achieved those 2 steps. 

The result is then plotted as a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r Average_Daily_Activity_Pattern}
byInterval <- tapply(Data$steps, Data$interval, mean, na.rm=TRUE)

plot(names(byInterval), byInterval, type='l', xlab="5-minute interval", ylab = "Average Steps Taken across All Days")
title(main="Average Daily Activity Pattern")
```

From plot, it can be seen that the interval at around 850 contains the maximum number of stages. We confirm this by running the code 
```{r maxInterval}
maxInterval <- names(byInterval)[which(byInterval == max(byInterval))]
```

The result is `r maxInterval`, having value of `r max(byInterval)`


## Imputing Missing Value

### Calulate the total number of missing values in the dataset

Using the code below, we calculate the number of missing value

```{r No of missing value}
noNA <- sum(is.na(Data$steps))
```

Thus, the total number of missing values in the dataset is `r noNA`.

### Filling in the missing values

The missing value is filled in using the average of the mean of the day and the mean of the 5-minutes interval. The value is given by the function below
```{r function to fill in missing value}
fillNa <- function(date, interval) {
    dayMean <- byDay[date] / sum(Data$date == date)
    intervalMean <- byInterval[as.character(interval)]
    if (dayMean != 0) {
        print("this")
        return((dayMean + intervalMean) / 2)
    } else {
        return(intervalMean)
    }
}
```

### Creating a new dataset without missing value

We first compute the `steps` for the new dataset. Then we create the new data steps using the computed `steps` and the subset od the old dataset.

```{r Fill in NA}
steps <- numeric()
for (i in 1:nrow(Data)) {
    if (is.na(Data$steps[i])) {
        steps <- c(steps, fillNa(as.character(Data$date[i]), Data$interval[i]))
    } else {
        steps <- c(steps, Data$steps[i])
    }
}
newData <- cbind(steps, Data[,2:3])
```

### Redrawn the histogram of the total number of steps taken each day with new dataset
```{r Histogram_for_total_steps_per_day_with_new_dataset}
library(ggplot2)
ggplot(newData) + geom_histogram(aes(x=date, weight=steps), bins=length(unique(newData$date))) + 
    labs(x="Date", y="No of Steps", title = "Histogram of The Total Number of Steps Taken Each Day(Completed)")
```

Finally, the new mean and median of total number of steps taken per day is calculated using functions as shown below: 

```{r new Mean and Median}
byDayNew <- tapply(newData$steps, newData$date, sum, na.rm=TRUE)
meanNew <- mean(byDayNew)
medianNew <- median(byDayNew)
```

Thus, we founded that the new mean is `r meanNew` and the new median is `r medianNew`.  Note that both mean and median have increased as a result of imputing the data.

## Differencrs in Activity Patterns Between Weekdays and Weekends

A new factor variable is first added into the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r Add a New Factor Variable}
newData$day <- sapply(newData$date, function(x){
    if (weekdays(x) %in% c("Saturday", "Sunday")) {
        return("weekend")
    } else {
        return("weekday")
    }
})
newData$day <- as.factor(newData$day)
```

A panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis) is then created.
```{r panel_plot}
library(dplyr)
library(ggplot2)
summary <- newData %>% group_by(day, interval) %>% summarize(avg=mean(steps))
ggplot(summary) + geom_line(aes(interval,avg)) + facet_grid(day~.) + labs(x="Interval") + 
    labs(y="Number of steps")
```
